Review Paper Draft

### 2020-06-29 lmm-genetic-review.tex
2.2 edits

\begin{abstract}
Many genetic studies that aim to identify genetic variants associated with complex phenotypes are subject to \anna{unobserved confounding factors, such as environmental heterogeneity.} This poses a challenge to the detection of multivariate associations of interest and is known to induce spurious associations when left unaccounted for. Penalized linear mixed models (LMMs) are an attractive method to correct for unobserved confounding. These methods correct for varying levels of relatedness and population structure by modeling it as a random effect with a covariance structure estimated from observed genetic data. \anna{Despite extensive literature on penalized regression, and LMMs, there is a need for a statistically rigorous review of penalized LMMs. Specifically, the ability of LMMs to accurately estimate fixed, sparse genetic effects in the presence of environmental heterogeneity, for which population structure may serve as a proxy, has not been well studied. Additionally, the non-genetic mechanism by which population stratification may confound the phenotype-genotype relationship is rarely discussed or treated explicitly in the statistical genetics literature.} Considering this, we evaluate the performance of several penalized regression methods in terms of MSE in the presence of a number of confounding structures.
% in the presence of confounding due to varying levels of environmental heterogeneity and relatedness.
\end{abstract}

>>>>

\begin{abstract}
Many genetic studies that aim to identify genetic variants associated with complex phenotypes are subject to unobserved confounding factors, such as environmental heterogeneity. This poses a challenge to the detection of multivariate associations of interest and is known to induce spurious associations when left unaccounted for. Penalized linear mixed models (LMMs) are an attractive method to correct for unobserved confounding. These methods correct for varying levels of relatedness and population structure by modeling it as a random effect with a covariance structure estimated from observed genetic data. Despite extensive literature on penalized regression, and LMMs, there is a need for a statistically rigorous review of penalized LMMs. Specifically, the ability of LMMs to accurately estimate fixed, sparse genetic effects in the presence of environmental heterogeneity, for which population structure may serve as a proxy, has not been well studied. Additionally, the non-genetic mechanism by which population stratification may confound the phenotype-genotype relationship is rarely discussed or treated explicitly in the statistical genetics literature. Considering this, we evaluate the performance of several penalized regression methods in terms of MSE in the presence of a number of confounding structures.
% in the presence of confounding due to varying levels of environmental heterogeneity and relatedness.
\end{abstract}

### 2020-06-29 main.tex
2.2 edits

Relatedness describes how genetically similar individuals are, and can refer to both ancient and recent relatedness.  Recent relatedness is characterized by small clusters of high levels of similarity and is indicative of familial or cryptic relatedness. Ancient relatedness, or common ancestry, is generally characterized by large clusters of low levels of similarity. The presence of multiple distinct ancestry groups is known as population stratification. \pb{Should some of these terms, like cryptic relatedness, be bolded?}

@@

The mechanism of environmental confounding is summarized in the directed acyclic graph displayed in figure \ref{fig:ps_env}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors \anna{directly} influence population-specific allele frequencies, or vice versa, only that the two may be related.  \anna{often end up related for complex reasons... the environment and genetics have become associated over time for complicated reasons that we aren't going to get into... air quality isn't changing people's genetics, genetics aren't causing people to move to areas of better/worse air quality}

@@

An additional consideration in genetic studies that we will briefly mention is that of genetic background. Genetic background describes when SNPs that have a causal relationship with the phenotype of interest are not directly observed, but are correlated with SNPs that are observed. As \cite{vilhjalmsson2012nature} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. While we acknowledge the challenges presented by genetic background, this work will focus on correction methods for confounding due to environmental heterogeneity, for which genetic similarity may serve as a proxy.\pb{Overall comment about this paragraph: I feel like we *are* addressing genetic background, by conducting a genomewide multivariate analysis.  I think its worth pointing out that this is not the only way to deal with genetic background, but it's not as though we're not addressing it at all.}

>>>>

Relatedness describes how genetically similar individuals are, and can refer to both ancient and recent relatedness.  Recent relatedness is characterized by small clusters of high levels of similarity and is indicative of familial or \textbf{cryptic relatedness}. Ancient relatedness, or common ancestry, is generally characterized by large clusters of low levels of similarity. The presence of multiple distinct ancestry groups is known as \textbf{population stratification}.

@@

The mechanism of environmental confounding is summarized in the directed acyclic graph displayed in figure \ref{fig:ps_env}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors directly influence population-specific allele frequencies, or vice versa, only that the two may be related. In the context of our air quality example, this means that we assume air quality is not directly altering individuals' allele frequencies, nor are allele frequencies causing individuals to move to regions of better or worse air quality. Genetic relatedness and environmental effects become associated over time for complex reasons, which are beyond the scope of this paper. As we will discuss, many methods of correcting for population structure rely on this relationship and do not formally differentiate between effects of population structure and those of environment. Such methods assume that genetic relatedness serves as an accurate proxy for environmental exposures, which may or may not be the case depending on the exposure in question.

@@

An additional consideration in genetic studies that we will briefly mention is that of genetic background. Genetic background describes when SNPs that have a causal relationship with the phenotype of interest are not directly observed, but are correlated with SNPs that are observed. As \cite{vilhjalmsson2012nature} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. Although there are a number of methods in which to account for genetic background \anna{what are they?}, the methods upon which we will focus address this by conducting a genomewide multivariate analysis.
